
---
title: "Predicting NFL Team Wins Using Linear Regression"
author: "Chuol Gatkek & Andrew Dingman"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This project applies linear regression to predict the number of wins for NFL teams using offensive and defensive statistics. Using data from 2018 to 2023, we explored how total points scored and allowed influence a team's success, with an emphasis on actionable insights for decision-makers.

## Load Libraries and Data

```{r}
# Load necessary libraries
library(tidyverse)
library(nflverse)
library(vroom)
```

## Prepare and Clean the Data

```{r}
# Load NFL game data from nflverse
games_data <- nflreadr::load_schedules()

# Filter regular season games from 2018 to 2023
games <- games_data %>%
  filter(season >= 2018 & season <= 2023, game_type == "REG") %>%
  select(season, week, team_home, team_away, result, home_score, away_score)

# Calculate wins per team
home_wins <- games %>%
  filter(result > 0) %>%
  group_by(team_home) %>%
  summarise(home_wins = n())

away_wins <- games %>%
  filter(result < 0) %>%
  group_by(team_away) %>%
  summarise(away_wins = n())

total_wins <- full_join(home_wins, away_wins, by = c("team_home" = "team_away")) %>%
  mutate(total_wins = coalesce(home_wins, 0) + coalesce(away_wins, 0)) %>%
  rename(team = team_home) %>%
  select(team, total_wins)
```

## Merge Wins with Points Scored and Allowed

```{r}
# Summarize offensive and defensive stats
team_stats <- games %>%
  mutate(
    pts_scored_home = home_score,
    pts_allowed_home = away_score,
    pts_scored_away = away_score,
    pts_allowed_away = home_score
  ) %>%
  select(team_home, pts_scored_home, pts_allowed_home, team_away, pts_scored_away, pts_allowed_away) %>%
  pivot_longer(cols = everything(), names_to = c("type", ".value"), names_pattern = "(.*)_(.*)") %>%
  rename(team = team, pts_scored = scored, pts_allowed = allowed) %>%
  group_by(team) %>%
  summarise(
    total_points_scored = sum(pts_scored, na.rm = TRUE),
    total_points_allowed = sum(pts_allowed, na.rm = TRUE)
  )

# Merge with win data
nfl_df <- left_join(total_wins, team_stats, by = "team")
```

## Build Linear Regression Model

```{r}
# Fit regression model
lm_model <- lm(total_wins ~ total_points_scored + total_points_allowed, data = nfl_df)
summary(lm_model)
```

## Visualize Model Results

```{r}
# Plot predicted vs actual
nfl_df$predicted_wins <- predict(lm_model)

ggplot(nfl_df, aes(x = total_wins, y = predicted_wins)) +
  geom_point(color = "steelblue", size = 3) +
  geom_abline(linetype = "dashed", color = "red") +
  labs(title = "Actual vs Predicted Wins", x = "Actual Wins", y = "Predicted Wins") +
  theme_minimal()
```

## Conclusion

This regression model revealed strong predictive power between scoring performance and win totals. Teams that score more and allow fewer points predictably win more games. This type of model can assist NFL analysts, coaches, and front offices in measuring team efficiency and identifying performance gaps.

